---
title: "SPI Data Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPI Data Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette walks through the main workflows supported by `spiR`: retrieving
SPI output data from GitHub, filtering to specific countries, years, pillars,
and dimensions, comparing statistical performance across countries, and
visualizing results with ggplot2.

## Getting Started

Install `spiR` from GitHub:

```{r install}
# Using remotes
remotes::install_github("WB-DECIS/spiR")

# Or using pak
pak::pak("WB-DECIS/spiR")
```

Then load the package:

```{r load}
library(spiR)
library(data.table)
library(ggplot2)
```

### The core data access functions

`spiR` provides access to five main SPI value-access functions, each returned as a
`data.table`:

| Function | Dataset | Format |
|---|---|---|
| `spi_data()` | Individual indicator scores | Wide — one row per country-year |
| `spi_index()` | Pillar and overall SPI scores | Wide — one row per country-year |
| `spi_aggregates()` | Aggregate/group scores | Long — one row per aggregate-year-indicator |
| `spi_indicator()` | Selected named indicator columns | Wide — one row per country-year |
| `country_info()` | Country metadata from `SPI_data.csv` | Wide — one row per country-year |

```{r datasets}
# Individual indicator scores for all countries and years
dt_data <- spi_data()

# Pillar-level and overall SPI index scores
dt_index <- spi_index()

# Aggregate/group scores (non-country rows only)
dt_agg <- spi_aggregates()

# Country-year metadata only
dt_country <- country_info()
```

Each dataset is a `data.table`. Key columns in `spi_index()`:

- `iso3c` — ISO 3166-1 alpha-3 country code
- `country` — country name
- `date` — year (integer)
- `SPI.INDEX` — overall SPI score (0–100)
- `SPI.INDEX.PIL1` through `SPI.INDEX.PIL5` — pillar scores

### Available versions

SPI data versions correspond to Git branches in the
[World Bank SPI repository](https://github.com/worldbank/SPI). Use
`spi_versions()` to see what is available:

```{r versions}
spi_versions()
#> [1] "master"   "SPI2024"  "SPI2023"  "SPI2022"  ...
```

The default version is `"master"` (latest stable). Pass any branch name as
the `version` argument to retrieve historical data.

---

## Filtering Data

The main retrieval functions share a common filter interface, with a few
type-specific arguments:

| Argument | Type | Description |
|---|---|---|
| `country` | Character vector of ISO3C codes | Applies to `data`, `index`, and `country_info()` |
| `region` | Character vector of region names | Applies to `aggregates` only |
| `year` | Integer or numeric vector | Applies to all types |
| `pillar` | Integer 1–5 | Column filter for `data`/`index`; row filter for `aggregates` |
| `dimension` | Character (e.g. `"5.2"`) | Column filter for `data`/`index`; row filter for `aggregates`. Overrides `pillar` when both supplied |
| `version` | Character branch name | Applies to all types |

### By country

```{r filter-country}
# Several European countries
europe <- spi_index(country = c("NOR", "SWE", "FIN", "DNK", "ISL"))
```

### By year

```{r filter-year}
# Most recent three years
recent <- spi_index(year = 2022:2024)

# Single year
y2024 <- spi_index(year = 2024L)
```

### By pillar

Restrict to indicator columns for a single pillar. Identifier columns
(`iso3c`, `date`, `country`, `SPI.INDEX`, etc.) are always returned.

```{r filter-pillar}
# Pillar 3: Data Products — only SDG-reporting indicator columns
dt_p3 <- spi_data(pillar = 3L)

# Pillar 5 index scores for all countries
idx_p5 <- spi_index(pillar = 5L)
```

### By dimension

Dimensions are coded as `"<pillar>.<dimension>"` (e.g. `"4.1"` for censuses
and surveys, `"5.2"` for standards and methods). When `dimension` is supplied
it overrides `pillar`.

```{r filter-dimension}
# Standards and methods (Pillar 5, Dimension 2)
standards <- spi_index(dimension = "5.2")

# Census and survey availability indicators only
census_ind <- spi_data(dimension = "4.1")
```

### Named indicators with `spi_indicator()`

If you already know the exact SPI indicator codes, `spi_indicator()` lets you
retrieve only those columns for the selected country/year range rather than the
full dimension. The first argument is `indicator`, so you can call it either
positionally or with `indicator =`.

```{r filter-indicator}
# One specific known indicator for Chile in 2024
spi_indicator(indicator = "SPI.D1.5.POV", country = "CHL", year = 2024L)

# Two known indicators, including raw values when available
spi_indicator(
  indicator = c("SPI.D1.5.POV", "SPI.D2.1.GDDS"),
  include_raw = TRUE
)
```

Setting `include_raw = TRUE` also returns the corresponding `RAW.D...`
columns when they are available in `SPI_data.csv`.

`spi_indicator()` is a filtered form of `spi_data()` that returns only the
requested indicator columns. Use `spi_data(dimension = ...)` when you want all
indicators in a dimension, and use `spi_indicator()` when you want a small set
of exact indicator columns.

### Country metadata with `country_info()`

Use `country_info()` when you want a clean metadata lookup table rather than
indicator values. It returns one row per country-year and keeps fields such as
ISO codes, region, income level, lending type, and population.

```{r filter-country-info}
# Metadata for Chile in 2024
chl_meta <- country_info(country = "CHL", year = 2024L)

# Metadata for a small country set across years
nordics_meta <- country_info(country = c("NOR", "SWE", "FIN"), year = 2022:2024)
```

### Metadata catalog access

When you need to inspect the SPI structure (pillars, dimensions, indicators)
before pulling values, use the metadata access functions.

The examples below use `eval = FALSE` so vignette builds do not require live
network calls.

```{r metadata-access, eval=FALSE}
  # Pillar-level metadata
  p <- metadata_pillars()

  # Dimensions for one pillar
  d <- metadata_dimensions(pillar = "2")

  # The metadata helpers also accept SPI metadata IDs
  d_id <- metadata_dimensions(pillar = "SPI.INDEX.PIL2")

  # Full associated metadata from one filter
  m <- metadata(pillar = "1")

  # Dimension filters can use either the canonical P.D value or the SPI ID
  m_dim <- metadata(dimension = "1.5")
  m_dim_id <- metadata(dimension = "SPI.DIM1.5.INDEX")

  # Indicator-focused metadata lookup
  i <- metadata(indicator = "SPI.D1.5.POV")
```

### Combining filters

Filters can be combined. Results are subsetted in all dimensions simultaneously:

```{r filter-combined}
# Sub-Saharan Africa members, Pillar 4, 2020–2024
ssa_p4 <- spi_index(
  country = c("KEN", "ETH", "TZA", "UGA", "RWA", "MOZ", "GHA", "SEN"),
  pillar  = 4L,
  year    = 2020:2024
)
```

### Regional aggregates

`spi_aggregates()` uses `region` instead of `country`. Region names follow
the World Bank classification:

```{r filter-aggregates}
# Pillar 1 scores for two regions
ssa_agg <- spi_aggregates(
  region = c("Africa Eastern and Southern", "Africa Western and Central"),
  pillar = 1L,
  year   = 2020:2024
)
```

### Using a specific version

```{r filter-version}
# Retrieve 2023 release data
dt_2023rel <- spi_index(version = "SPI2023")
```

---

## Comparing Countries

### Pillar profiles for a set of countries

Retrieve index scores, reshape to long format with `data.table::melt()`, and
compare pillar profiles across countries:

```{r compare-pillars}
# Select a diverse set of countries
countries <- c("NOR", "MEX", "KEN", "BGD", "UGA", "BRA", "IND", "ETH")

idx <- spi_index(country = countries, year = 2024L)

# Keep only the five pillar score columns plus identifiers
pillar_cols <- c("country", "iso3c", "date",
                 paste0("SPI.INDEX.PIL", 1:5))
idx_pillars <- idx[, ..pillar_cols]

# Reshape wide → long: one row per country-pillar
idx_long <- melt(
  idx_pillars,
  id.vars         = c("country", "iso3c", "date"),
  measure.vars    = paste0("SPI.INDEX.PIL", 1:5),
  variable.name   = "pillar",
  value.name      = "score",
  variable.factor = FALSE
)

# Map pillar codes to readable labels
pillar_labels <- setNames(
  paste("Pillar", 1:5),
  paste0("SPI.INDEX.PIL", 1:5)
)
idx_long[, pillar_label := pillar_labels[pillar]]
```

### Tracking a country over time

```{r time-trend}
# Overall SPI index score for Kenya, 2016–2024
ken <- spi_index(country = "KEN")
ken_trend <- ken[, .(date, SPI.INDEX,
                     SPI.INDEX.PIL1, SPI.INDEX.PIL2, SPI.INDEX.PIL3,
                     SPI.INDEX.PIL4, SPI.INDEX.PIL5)]

# Reshape to long for plotting all pillars over time
ken_long <- melt(
  ken_trend,
  id.vars       = "date",
  variable.name = "series",
  value.name    = "score"
)
```

### Top and bottom performers

```{r top-bottom}
# All countries in the latest year
latest_year <- spi_index(year = 2024L)[
  , .(country, iso3c, SPI.INDEX,
      SPI.INDEX.PIL1, SPI.INDEX.PIL2, SPI.INDEX.PIL3,
      SPI.INDEX.PIL4, SPI.INDEX.PIL5)
]

# Sort by overall score
setorder(latest_year, -SPI.INDEX)

# Top 10 and bottom 10
top10    <- head(latest_year, 10)
bottom10 <- tail(latest_year, 10)
```

---

## Visualization

The examples below use `ggplot2`.

### Overall SPI score by country (bar chart)

```{r plot-bar}
# 8 countries for a compact comparison
countries <- c("NOR", "MEX", "KEN", "BGD", "UGA", "BRA", "IND", "ETH")
idx <- spi_index(country = countries, year = 2024L)

# Exclude countries with no overall score in this year
idx <- idx[!is.na(SPI.INDEX)]

ggplot(idx, aes(x = reorder(country, SPI.INDEX), y = SPI.INDEX)) +
  geom_col(fill = "#0071BC") +
  geom_text(aes(label = round(SPI.INDEX, 1)),
            hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = 0)) +
  labs(
    title   = "SPI Overall Score (2024)",
    x       = NULL,
    y       = "SPI Score (0–100)",
    caption = "Source: World Bank Statistical Performance Indicators"
  ) +
  theme_minimal(base_size = 12)
```

### Pillar breakdown across countries (grouped bar)

```{r plot-grouped-bar}
pillar_cols  <- paste0("SPI.INDEX.PIL", 1:5)
idx_pillars  <- idx[, c("country", pillar_cols), with = FALSE]

idx_long <- melt(
  idx_pillars,
  id.vars         = "country",
  measure.vars    = pillar_cols,
  variable.name   = "pillar",
  value.name      = "score",
  variable.factor = FALSE
)

pillar_labels <- setNames(
  paste("Pillar", 1:5),
  paste0("SPI.INDEX.PIL", 1:5)
)
idx_long[, pillar_label := pillar_labels[pillar]]

ggplot(idx_long, aes(x = country, y = score, fill = pillar_label)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = 0)) +
  labs(
    title   = "SPI Pillar Scores by Country (2024)",
    x       = NULL,
    y       = "Score (0–100)",
    caption = "Source: World Bank Statistical Performance Indicators"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

### Time trend (line chart)

```{r plot-trend}
# Overall SPI score over time for a group of countries
countries <- c("NOR", "KEN", "BGD", "ETH")
idx       <- spi_index(country = countries)

ggplot(idx, aes(x = date, y = SPI.INDEX, colour = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(2016, 2024, by = 2)) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title   = "SPI Overall Score Over Time",
    x       = NULL,
    y       = "SPI Score (0–100)",
    colour  = NULL,
    caption = "Source: World Bank Statistical Performance Indicators"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Pillar radar / heatmap (faceted bar)

A faceted view makes it easy to compare pillar strengths and weaknesses
across countries:

```{r plot-facet}
countries <- c("NOR", "MEX", "KEN", "BGD")
idx       <- spi_index(country = countries, year = 2024L)

pillar_cols <- paste0("SPI.INDEX.PIL", 1:5)
idx_long    <- melt(
  idx[, c("country", pillar_cols), with = FALSE],
  id.vars         = "country",
  measure.vars    = pillar_cols,
  variable.name   = "pillar",
  value.name      = "score",
  variable.factor = FALSE
)
idx_long[, pillar_label := paste("Pillar", sub("SPI.INDEX.PIL", "", pillar, fixed = TRUE))]

ggplot(idx_long, aes(x = pillar_label, y = score, fill = pillar_label)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey60") +
  facet_wrap(~country, ncol = 2) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = 0)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title   = "Pillar Scores by Country (2024)",
    x       = NULL,
    y       = "Score (0–100)",
    caption = "Source: World Bank Statistical Performance Indicators"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        strip.text  = element_text(face = "bold"))
```

### Regional aggregates over time

```{r plot-aggregates}
regions <- c(
  "Africa Eastern and Southern",
  "Africa Western and Central",
  "South Asia",
  "Europe & Central Asia"
)

agg <- spi_aggregates(region = regions, year = 2019:2024)

# Filter to overall SPI index scores only
agg_spi <- agg[source_id == "SPI.INDEX"]
# Use a descriptive column name for the legend
agg_spi[, region_label := country]

ggplot(agg_spi, aes(x = date, y = value, colour = region_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    title   = "SPI Overall Score by Region",
    x       = NULL,
    y       = "SPI Score (0–100)",
    colour  = "Region",
    caption = "Source: World Bank Statistical Performance Indicators"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text      = element_text(size = 9)) +
  guides(colour = guide_legend(ncol = 2))
```

---

## Caching and Versions

`spiR` caches downloaded files in-session to avoid redundant network requests.
Downloaded files are cached per data type and version, so repeated calls with
different `country` or `year` filters draw from the same local cache rather
than re-downloading:

```{r cache}
# First call: downloads from GitHub
dt <- spi_index(year = 2024L)

# Subsequent calls: served from cache instantly (different filter, same file)
dt <- spi_index(year = 2022L)

# Clear the cache (e.g., after a data update)
spi_clear_cache()

# Force a fresh download (cache was cleared)
dt <- spi_index(year = 2024L)
```

### Managing the data inventory

`spiR` also maintains a local inventory of available SPI data versions
(GitHub branches). The inventory is fetched automatically on first use and
cached between sessions. Use these functions to manage it:

```{r inventory}
# Refresh the inventory from GitHub (e.g., after a new SPI release)
spi_update_inventory()

# Remove the local inventory cache entirely (forces a fresh fetch on next call)
spi_clear_inventory()
```

To compare data across SPI releases, use the `version` argument:

```{r compare-versions}
idx_current <- spi_index(country = "KEN")
idx_2023rel <- spi_index(country = "KEN", version = "SPI2023")
idx_2022rel <- spi_index(country = "KEN", version = "SPI2022")
```

---

## Further Reading

- [SPI Program page](https://www.worldbank.org/en/programs/statistical-performance-indicators) — overview, interactive data explorer
- [SPI Handbook](https://worldbank.github.io/SPI/measuring-the-statistical-performance-of-countries-an-overview-of-the-statistical-performance-indicators-and-index.html#indicator-metadata) — full methodology and indicator metadata
- [SPI GitHub repository](https://github.com/worldbank/SPI) — raw data and code
- [spiR function reference](https://wb-decis.github.io/spiR/reference/) — all exported functions
